---
title: "ilant_health_take-home_project"
author: "Jim Tinley"
date: "2023-08-26"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Importing libraries
```{r, message=FALSE}
library(data.table)
library(tidyr) 
library(ggplot2)
library(readr)
library(knitr)
library(corrplot)
library(psych)
library(cobalt)
library(caret)
library(pscl)
library(rms)
library(fastDummies)
library(mltools)
library(stringr)
library(kableExtra)
library(car)
library(tidyverse) 
library(Matching)
library(WeightIt)
library(gamlr)
library(glmnet)
``` 

```{r}
# setting working directory
setwd("~/Documents/ilant_project")
```

# Importing files

## Beneficiary summary files - main file with patient features
```{r}
ben_1_08_summary_dt <- fread("DE1_0_2008_Beneficiary_Summary_File_Sample_20.csv")
ben_1_09_summary_dt <- fread("DE1_0_2009_Beneficiary_Summary_File_Sample_20.csv")
ben_1_10_summary_dt <- fread("DE1_0_2010_Beneficiary_Summary_File_Sample_20.csv")

ben_2_08_summary_dt <- fread("DE1_0_2008_Beneficiary_Summary_File_Sample_2.csv")
ben_2_09_summary_dt <- fread("DE1_0_2009_Beneficiary_Summary_File_Sample_2.csv")
ben_2_10_summary_dt <- fread("DE1_0_2010_Beneficiary_Summary_File_Sample_2.csv")

ben_3_08_summary_dt <- fread("DE1_0_2008_Beneficiary_Summary_File_Sample_3.csv")
ben_3_09_summary_dt <- fread("DE1_0_2009_Beneficiary_Summary_File_Sample_3.csv")
ben_3_10_summary_dt <- fread("DE1_0_2010_Beneficiary_Summary_File_Sample_3.csv")

ben_4_08_summary_dt <- fread("DE1_0_2008_Beneficiary_Summary_File_Sample_4.csv")
ben_4_09_summary_dt <- fread("DE1_0_2009_Beneficiary_Summary_File_Sample_4.csv")
ben_4_10_summary_dt <- fread("DE1_0_2010_Beneficiary_Summary_File_Sample_4.csv")

ben_5_08_summary_dt <- fread("DE1_0_2008_Beneficiary_Summary_File_Sample_5.csv")
ben_5_09_summary_dt <- fread("DE1_0_2009_Beneficiary_Summary_File_Sample_5.csv")
ben_5_10_summary_dt <- fread("DE1_0_2010_Beneficiary_Summary_File_Sample_5.csv")

ben_6_08_summary_dt <- fread("DE1_0_2008_Beneficiary_Summary_File_Sample_6.csv")
ben_6_09_summary_dt <- fread("DE1_0_2009_Beneficiary_Summary_File_Sample_6.csv")
ben_6_10_summary_dt <- fread("DE1_0_2010_Beneficiary_Summary_File_Sample_6.csv")

ben_7_08_summary_dt <- fread("DE1_0_2008_Beneficiary_Summary_File_Sample_7.csv")
ben_7_09_summary_dt <- fread("DE1_0_2009_Beneficiary_Summary_File_Sample_7.csv")
ben_7_10_summary_dt <- fread("DE1_0_2010_Beneficiary_Summary_File_Sample_7.csv")
```

### Creating combined beneficiary summary files - 1/year
```{r}
beneficiary_08_summary_files <- list(
  ben_1_08_summary_dt, ben_2_08_summary_dt, ben_3_08_summary_dt, 
  ben_4_08_summary_dt, ben_5_08_summary_dt, ben_6_08_summary_dt, 
  ben_7_08_summary_dt
)

beneficiary_09_summary_files <- list(
  ben_1_09_summary_dt, ben_2_09_summary_dt, ben_3_09_summary_dt,
  ben_4_09_summary_dt, ben_5_09_summary_dt,ben_6_09_summary_dt, 
  ben_7_09_summary_dt
)

beneficiary_10_summary_files <- list(
  ben_1_10_summary_dt, ben_2_10_summary_dt, ben_3_10_summary_dt,
  ben_4_10_summary_dt, ben_5_10_summary_dt, ben_6_10_summary_dt,
  ben_7_10_summary_dt
)

ben_08_summary_DT <- rbindlist(
  beneficiary_08_summary_files,
  use.names = TRUE,
  fill = TRUE,
  idcol = "file"
)

ben_09_summary_DT <- rbindlist(
  beneficiary_09_summary_files,
  use.names = TRUE,
  fill = TRUE,
  idcol = "file"
)

ben_10_summary_DT <- rbindlist(
  beneficiary_10_summary_files,
  use.names = TRUE,
  fill = TRUE,
  idcol = "file"
)

# Removing intermediate files
rm(ben_1_08_summary_dt, ben_1_09_summary_dt, ben_1_10_summary_dt,
  ben_2_08_summary_dt, ben_2_09_summary_dt, ben_2_10_summary_dt,
  ben_3_08_summary_dt, ben_3_09_summary_dt, ben_3_10_summary_dt,
  ben_4_08_summary_dt, ben_4_09_summary_dt, ben_4_10_summary_dt,
  ben_5_08_summary_dt, ben_5_09_summary_dt, ben_5_10_summary_dt,
  ben_6_08_summary_dt, ben_6_09_summary_dt, ben_6_10_summary_dt,
  ben_7_08_summary_dt, ben_7_09_summary_dt, ben_7_10_summary_dt, beneficiary_08_summary_files, beneficiary_09_summary_files, beneficiary_10_summary_files)
```


## Prescription Drug Events - did not use in analysis
```{r}
# Not using this drug data in analysis
# drug_1_dt <- fread("DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_20.csv")
# drug_2_dt <- fread("DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_2.csv")
# drug_3_dt <- fread("DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_3.csv")
# drug_4_dt <- fread("DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_4.csv")
# drug_5_dt <- fread("DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_5.csv")
# drug_6_dt <- fread("DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_6.csv")
# drug_7_dt <- fread("DE1_0_2008_to_2010_Prescription_Drug_Events_Sample_7.csv")
```

```{r}
# drug_files <- list(
#   drug_1_dt, drug_2_dt, drug_3_dt, 
#   drug_4_dt, drug_5_dt, drug_6_dt, 
#   drug_7_dt)
# 
# drug_DT <- rbindlist(
#   drug_files,  
#   use.names = TRUE, 
#   fill = TRUE,
#   idcol ="file")
# 
# rm(drug_1_dt, drug_2_dt, drug_3_dt, 
#   drug_4_dt, drug_5_dt, drug_6_dt, 
#   drug_7_dt, drug_files)
```

## Carrier claims files
```{r}
# Importing
carrier_claims_1a_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_20A.csv")
#carrier_claims_1b_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_20B.csv")

carrier_claims_2a_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_2A.csv")
#carrier_claims_2b_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_2B.csv")

carrier_claims_3a_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_3A.csv")
#carrier_claims_3b_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_3B.csv")

carrier_claims_4a_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_4A.csv")
#carrier_claims_4b_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_4B.csv")

carrier_claims_5a_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_5A.csv")
#carrier_claims_5b_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_5B.csv")

carrier_claims_6a_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_6A.csv")
#carrier_claims_6b_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_6B.csv")

carrier_claims_7a_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_7A.csv")
#carrier_claims_7b_dt <-fread("DE1_0_2008_to_2010_Carrier_Claims_Sample_7B.csv")
```

### Custom function: filter claims files on BMI and indicate obesity & ER_visits
```{r}
BMI_ER_filter_function <- function(input_dt) {
  # Create has BMI column
  icd9_columns <- grep("^ICD9_DGNS_CD_", names(input_dt), value = TRUE)
  input_dt[, has_BMI := 0]
  for (col in icd9_columns) {
    input_dt[, has_BMI := has_BMI | get(col) %like% "^V85"]
  }
  
  # Create obese column based on specific ICD-9 codes
  input_dt[, obese := 0]
  for (col in icd9_columns) {
    input_dt[, obese := obese | get(col) %like% "^V853" | get(col) %like% "^V854" | get(col) %like% "^278"]
  }
  input_dt[, obese := as.integer(obese)]
  
  # Filter for cases with has BMI
  input_dt <- input_dt[has_BMI == 1]
  
  # Create ER_visit column
  hcpcs_columns <- grep("^HCPCS_CD_", names(input_dt), value = TRUE)
  input_dt[, ER_visit := 0]
  for (col in hcpcs_columns) {
    input_dt[, ER_visit := ER_visit | get(col) %like% "^9928"]
  }

  
  return(input_dt)
}

```


### Using Custom function to pre-filter carrier claims files
```{r}
carrier_tables <- c(
  "carrier_claims_1a_dt",
  "carrier_claims_2a_dt", 
  "carrier_claims_3a_dt", 
  "carrier_claims_4a_dt",
  "carrier_claims_5a_dt",
  "carrier_claims_6a_dt", 
  "carrier_claims_7a_dt"
)

for (table_name in carrier_tables) {
  assign(table_name, BMI_ER_filter_function(get(table_name)))
}
```

### Combining filtered carrier claims files
```{r}
# Creating a combined carrier claims file
carrier_claims_files <- list(
  carrier_claims_1a_dt,
  carrier_claims_2a_dt,
  carrier_claims_3a_dt,
  carrier_claims_4a_dt,
  carrier_claims_5a_dt,
  carrier_claims_6a_dt,
  carrier_claims_7a_dt
)

carrier_claims_DT <- rbindlist(
  carrier_claims_files,
  use.names = TRUE, 
  fill = TRUE,
  idcol ="file")

rm(carrier_claims_1a_dt, carrier_claims_1b_dt,
  carrier_claims_2a_dt, carrier_claims_2b_dt,
  carrier_claims_3a_dt, carrier_claims_3b_dt,
  carrier_claims_4a_dt, carrier_claims_4b_dt,
  carrier_claims_5a_dt, carrier_claims_5b_dt,
  carrier_claims_6a_dt, carrier_claims_6b_dt,
  carrier_claims_7a_dt, carrier_claims_7b_dt, carrier_claims_files
)
```

```{r}
# Checking for unique patients within data
length(unique(carrier_claims_DT$DESYNPUF_ID))
```

### Splitting carrier claims into repsective years (needed for later joins)
```{r}
# splitting carrier claims into three year periods to ensure accuracy of joins with ben_summary
carrier_08_claims_DT <- carrier_claims_DT[CLM_FROM_DT %like% "^2008"]
carrier_09_claims_DT <- carrier_claims_DT[CLM_FROM_DT %like% "^2009"]
carrier_10_claims_DT <- carrier_claims_DT[CLM_FROM_DT %like% "^2010"]
```


## Outpatient claims files
```{r}
# Importing
outpatient_1_claims_dt <- fread("DE1_0_2008_to_2010_Outpatient_Claims_Sample_20.csv")
outpatient_2_claims_dt <- fread("DE1_0_2008_to_2010_Outpatient_Claims_Sample_2.csv")
outpatient_3_claims_dt <- fread("DE1_0_2008_to_2010_Outpatient_Claims_Sample_3.csv")
outpatient_4_claims_dt <- fread("DE1_0_2008_to_2010_Outpatient_Claims_Sample_4.csv")
outpatient_5_claims_dt <- fread("DE1_0_2008_to_2010_Outpatient_Claims_Sample_5.csv")
outpatient_6_claims_dt <- fread("DE1_0_2008_to_2010_Outpatient_Claims_Sample_6.csv")
outpatient_7_claims_dt <- fread("DE1_0_2008_to_2010_Outpatient_Claims_Sample_7.csv")
```

### Using Custom function to pre-filter outpatient claims files
```{r}
outpatient_tables <- c(
  "outpatient_1_claims_dt",
  "outpatient_2_claims_dt", 
  "outpatient_3_claims_dt",
  "outpatient_4_claims_dt",
  "outpatient_5_claims_dt",
  "outpatient_6_claims_dt", 
  "outpatient_7_claims_dt"
)

for (table_name in outpatient_tables) {
  assign(table_name, BMI_ER_filter_function(get(table_name)))
}

```

### Combining filtered outpatient claims files
```{r}
outpatient_claims_files <- list(
  outpatient_1_claims_dt, outpatient_2_claims_dt,
  outpatient_3_claims_dt, outpatient_4_claims_dt,
  outpatient_5_claims_dt, outpatient_6_claims_dt,
  outpatient_7_claims_dt
)

outpatient_claims_DT <- rbindlist(
  outpatient_claims_files,
  use.names = TRUE, 
  fill = TRUE,
  idcol ="file")

rm(outpatient_1_claims_dt, outpatient_2_claims_dt,
  outpatient_3_claims_dt, outpatient_4_claims_dt,
  outpatient_5_claims_dt, outpatient_6_claims_dt,
  outpatient_7_claims_dt, outpatient_claims_files
)
```

```{r}
length(unique(outpatient_claims_DT$DESYNPUF_ID))
```

### Splitting outpatient claims into repsective years (needed for later joins)
```{r}
# splitting carrier claims into three year periods to ensure accuracy of joins with ben_summary
outpatient_08_claims_DT <- outpatient_claims_DT[CLM_FROM_DT %like% "^2008"]
outpatient_09_claims_DT <- outpatient_claims_DT[CLM_FROM_DT %like% "^2009"]
outpatient_10_claims_DT <- outpatient_claims_DT[CLM_FROM_DT %like% "^2010"]
```


## Inpatient Claims (excluded from analysis) - missing data all revenue center data (HCPCS) needed for ER coding
```{r}

# inpatient_1_claims_dt <-fread("DE1_0_2008_to_2010_Inpatient_Claims_Sample_20.csv")
# inpatient_2_claims_dt <-fread("DE1_0_2008_to_2010_Inpatient_Claims_Sample_2.csv")
# inpatient_3_claims_dt <-fread("DE1_0_2008_to_2010_Inpatient_Claims_Sample_3.csv")
# inpatient_4_claims_dt <-fread("DE1_0_2008_to_2010_Inpatient_Claims_Sample_4.csv")
# inpatient_5_claims_dt <-fread("DE1_0_2008_to_2010_Inpatient_Claims_Sample_5.csv")
# inpatient_6_claims_dt <-fread("DE1_0_2008_to_2010_Inpatient_Claims_Sample_6.csv")
# inpatient_7_claims_dt <-fread("DE1_0_2008_to_2010_Inpatient_Claims_Sample_7.csv")
```

```{r}
# SKIP HCPCS data is missing for all files--cannot determine ER visit
# inpatient_1_claims_dt <- BMI_ER_filter_function(inpatient_1_claims_dt)
# inpatient_2_claims_dt <- BMI_ER_filter_function(inpatient_2_claims_dt)
# inpatient_3_claims_dt <- BMI_ER_filter_function(inpatient_3_claims_dt)
# inpatient_4_claims_dt <- BMI_ER_filter_function(inpatient_4_claims_dt)
# inpatient_5_claims_dt <- BMI_ER_filter_function(inpatient_5_claims_dt)
# inpatient_6_claims_dt <- BMI_ER_filter_function(inpatient_6_claims_dt)
# inpatient_7_claims_dt <- BMI_ER_filter_function(inpatient_7_claims_dt)
```

# Joining data


## Joining 2008 outpatient and carrier claims data
```{r}
carrier_08_patient_IDs <- carrier_08_claims_DT[,c("DESYNPUF_ID","obese", "ER_visit")]
outpatient_08_patient_IDs <- outpatient_08_claims_DT[,c("DESYNPUF_ID", "obese", "ER_visit")]

patients_08_DT <- rbindlist(list(
  carrier_08_patient_IDs,
  outpatient_08_patient_IDs),
  use.names = TRUE,
  fill = TRUE,
  idcol = "file"
)

patients_08_DT <- patients_08_DT[!duplicated(DESYNPUF_ID)]
```

## Joining 2009 outpatient and carrier claims data
```{r}
carrier_09_patient_IDs <- carrier_09_claims_DT[,c("DESYNPUF_ID","obese", "ER_visit")]
outpatient_09_patient_IDs <- outpatient_09_claims_DT[,c("DESYNPUF_ID","obese", "ER_visit")]

patients_09_DT <- rbindlist(list(
  carrier_09_patient_IDs,
  outpatient_09_patient_IDs),
  use.names = TRUE,
  fill = TRUE,
  idcol = "file"
)

patients_09_DT <- patients_09_DT[!duplicated(DESYNPUF_ID)]
```

## Joining 2010 outpatient and carrier claims data
```{r}
carrier_10_patient_IDs <- carrier_10_claims_DT[,c("DESYNPUF_ID","obese", "ER_visit")]
outpatient_10_patient_IDs <- outpatient_10_claims_DT[,c("DESYNPUF_ID","obese", "ER_visit")]

patients_10_DT <- rbindlist(list(
  carrier_10_patient_IDs,
  outpatient_10_patient_IDs),
  use.names = TRUE,
  fill = TRUE,
  idcol = "file"
)

patients_10_DT <- patients_10_DT[!duplicated(DESYNPUF_ID)]
```


## Joining claims files based on unique patient ID with beneficiary summary
```{r}
ben_claims_08_merge <- merge(ben_08_summary_DT, patients_08_DT, 
                           by = "DESYNPUF_ID", all.y = TRUE)
```

```{r}
ben_claims_09_merge <- merge(ben_09_summary_DT, patients_09_DT, 
                           by = "DESYNPUF_ID", all.y = TRUE)
```

```{r}
ben_claims_10_merge <- merge(ben_10_summary_DT, patients_10_DT, 
                           by = "DESYNPUF_ID", all.y = TRUE)
```

```{r}
all_patients_summary_DT <- rbindlist(list(
  ben_claims_08_merge,
  ben_claims_09_merge,
  ben_claims_10_merge),
  use.names = TRUE,
  fill = TRUE,
  idcol = "file"
)

all_patients_summary_DT <- all_patients_summary_DT[, -c("file.x", "file.y")]

# YOY patient overlap
length(unique(all_patients_summary_DT$DESYNPUF_ID))

# removing YOY patient overlap
all_patients_summary_DT <- all_patients_summary_DT[!duplicated(DESYNPUF_ID)]
```

## Cleaning data for analysis
```{r}
# Adding age variable

reference_date <- as.Date("2011-01-01")
all_patients_summary_DT[, BENE_BIRTH_DT := as.Date(as.character(BENE_BIRTH_DT), format = "%Y%m%d")]
all_patients_summary_DT[, age := as.integer((reference_date - BENE_BIRTH_DT) / 365.25)]
all_patients_summary_DT <- all_patients_summary_DT[, -c("BENE_BIRTH_DT")]
```

```{r}
# Adding deceased binary

all_patients_summary_DT[, deceased := as.numeric(ifelse(BENE_DEATH_DT>0, 1,0))]
all_patients_summary_DT[is.na(deceased), deceased := 0]
all_patients_summary_DT <- all_patients_summary_DT[, -c("BENE_DEATH_DT")]
```

```{r}
# Creating 0/1 sex binary
all_patients_summary_DT[, female := as.numeric(ifelse(BENE_SEX_IDENT_CD==2, 1,0))]

# Dropping original categorical variable BENE_SEX
all_patients_summary_DT <- all_patients_summary_DT[, -c("BENE_SEX_IDENT_CD")]
```

```{r}
# One-hot encoding race; first creating categorical variable
all_patients_summary_DT[, race := 
  ifelse(BENE_RACE_CD == 1, "white",
         ifelse(BENE_RACE_CD == 2, "black",
                ifelse(BENE_RACE_CD == 3, "other",
                       ifelse(BENE_RACE_CD == 5, "hispanic", "unknown")
                )
         )
  )
]

# One-hot
all_patients_summary_DT$race <- as.factor(all_patients_summary_DT$race)
all_patients_summary_DT <- all_patients_summary_DT[,c("black","hispanic","other","white") := one_hot(as.data.table(all_patients_summary_DT$race))]

# Dropping original categorical variable and baseline (white) to avoid multicolinearity 
all_patients_summary_DT <- all_patients_summary_DT[, -c("white","race","BENE_RACE_CD")]
```

```{r}
# reformatting all chronic conditions 0/1 binaries
all_patients_summary_DT[, alzheimer_dementia := as.numeric(
  ifelse(SP_ALZHDMTA==1, 1,0))]

all_patients_summary_DT[, chronic_heart_failure := as.numeric(
  ifelse(SP_CHF==1, 1,0))]

all_patients_summary_DT[, kidney_disease := as.numeric(
  ifelse(SP_CHRNKIDN==1, 1,0))]

all_patients_summary_DT[, cancer := as.numeric(
  ifelse(SP_CNCR==1, 1,0))]

all_patients_summary_DT[, copd := as.numeric(
  ifelse(SP_COPD==1, 1,0))]

all_patients_summary_DT[, depression := as.numeric(
  ifelse(SP_DEPRESSN==1, 1,0))]

all_patients_summary_DT[, diabetes := as.numeric(
  ifelse(SP_DIABETES==1, 1,0))]

all_patients_summary_DT[, ischemic_heart_disease := as.numeric(
  ifelse(SP_ISCHMCHT==1, 1,0))]

all_patients_summary_DT[, osteoporosis := as.numeric(
  ifelse(SP_OSTEOPRS==1, 1,0))]

all_patients_summary_DT[, rheumatoid_and_osteoarthritis := as.numeric(
  ifelse(SP_RA_OA==1, 1,0))]

all_patients_summary_DT[, stroke_transient_ischemic_attack := as.numeric(
  ifelse(SP_STRKETIA==1, 1,0))]

# Creating binary for end stage renal disease
all_patients_summary_DT[, end_stage_renal_disease := as.numeric(
  ifelse(BENE_ESRD_IND=="Y", 1,0))]

# Creating health insurance binaries for PartA-D
all_patients_summary_DT[, part_A_coverage := as.numeric(
  ifelse(BENE_HI_CVRAGE_TOT_MONS>0, 1,0))]

all_patients_summary_DT[, part_B_coverage := as.numeric(
  ifelse(BENE_SMI_CVRAGE_TOT_MONS>0, 1,0))]

all_patients_summary_DT[, HMO_coverage := as.numeric(
  ifelse(BENE_HMO_CVRAGE_TOT_MONS>0, 1,0))]

all_patients_summary_DT[, part_D_coverage := as.numeric(
  ifelse(PLAN_CVRG_MOS_NUM>0, 1,0))]


# Dropping all redundant columns
all_patients_summary_DT <- all_patients_summary_DT[, -c(
"SP_ALZHDMTA","SP_CHF", "SP_CHRNKIDN",
"SP_CHF", "SP_CNCR", "SP_COPD",
"SP_DEPRESSN", "SP_DIABETES", 
"SP_ISCHMCHT", "SP_OSTEOPRS", 
"SP_RA_OA", "SP_STRKETIA", "BENE_ESRD_IND",
"BENE_HI_CVRAGE_TOT_MONS",
"BENE_SMI_CVRAGE_TOT_MONS",
"BENE_HMO_CVRAGE_TOT_MONS",
"PLAN_CVRG_MOS_NUM")]
```



## Using given SSA county codes to get supplemental data on Rurality (proxy for hospital access)
```{r}
# concatenate state/county ssa codes
all_patients_summary_DT[, ssa_state_co :=  paste0(SP_STATE_CODE, BENE_COUNTY_CD)]
```

```{r}
# removing those with unknown county codes or states
all_patients_summary_DT <- all_patients_summary_DT[BENE_COUNTY_CD>0 & BENE_COUNTY_CD<999 ]
all_patients_summary_DT <- all_patients_summary_DT[SP_STATE_CODE<54]
```

```{r}
# adding third digit to county code to match crosswalk to fips
all_patients_summary_DT[, BENE_COUNTY_CD := sprintf("%03d", BENE_COUNTY_CD)]
```

```{r}
# using CMS crosswalk to get fips codes
ssa_fips_crosswalk_dt <-fread("ssa_fips_CW.csv")

ssa_fips_crosswalk_dt <- ssa_fips_crosswalk_dt[, c("ssacounty", "fipscounty", "fipsstate")]
ssa_fips_crosswalk_dt$ssacounty <- as.character(ssa_fips_crosswalk_dt$ssacounty)
```

```{r}
# merging fips codes on to main patient file
setkey(all_patients_summary_DT, ssa_state_co)
setkey(ssa_fips_crosswalk_dt, ssacounty)

all_patients_summary_DT <- merge(all_patients_summary_DT, ssa_fips_crosswalk_dt, by.x = "ssa_state_co", by.y = "ssacounty", all.x = TRUE)
```

```{r}
# removing those without fips data
all_patients_summary_DT <- all_patients_summary_DT[!is.na(fipscounty)]
```

```{r}
# importing fips rucc crosswalk
fips_rucc_crosswalk_dt <-fread("ruralurbancodes2013.csv")
```

```{r}
# merging fips codes on to main patient file
setkey(all_patients_summary_DT, fipscounty)
setkey(fips_rucc_crosswalk_dt, FIPS)

all_patients_summary_DT <- merge(all_patients_summary_DT, fips_rucc_crosswalk_dt, by.x = "fipscounty", by.y = "FIPS", all.x = TRUE)

# resetting key back to unique patient id
setkey(all_patients_summary_DT, DESYNPUF_ID)
```

```{r}
# Reducing down to final columns for analysis
FINAL_ANALYSIS_DT <- all_patients_summary_DT[, -c(
  "fipscounty", "ssa_state_co", "file", "SP_STATE_CODE", "BENE_COUNTY_CD",  "MEDREIMB_IP", "MEDREIMB_OP", "MEDREIMB_OP", "MEDREIMB_CAR", "BENRES_CAR", "BENRES_IP", "BENRES_OP", "PPPYMT_IP", "PPPYMT_OP", "PPPYMT_CAR", "fipsstate", "State", "Population_2010", "County_Name", "RUCC_2013", "deceased")]
FINAL_ANALYSIS_DT$ER_visit <- as.integer(FINAL_ANALYSIS_DT$ER_visit)
```

```{r}
# One-hot encoding Metro description baseline will be most populous urban area
FINAL_ANALYSIS_DT$Description <- as.factor(FINAL_ANALYSIS_DT$Description)
metro_dt <- one_hot(as.data.table(FINAL_ANALYSIS_DT$Description))
```

```{r}
FINAL_ANALYSIS_DT <- FINAL_ANALYSIS_DT[,c(
  "Metro - 1 million pop or more",
  "Metro - 250,000 to 1 million pop",
  "Metro - fewer than 250,000 pop",
  "Nonmetro - rural-less than 2,500 urban pop, adjacent to a metro area",
  "Nonmetro - rural-less than 2,500 urban pop, not adjacent to a metro area",
  "Nonmetro - Urban pop of 2,500 to 19,999, adjacent to a metro area" ,
  "Nonmetro - Urban pop of 2,500 to 19,999, not adjacent to a metro area",
  "Nonmetro - Urban pop of 20,000 or more, adjacent to a metro area", 
  "Nonmetro - Urban pop of 20,000 or more, not adjacent to a metro area") := one_hot(as.data.table(FINAL_ANALYSIS_DT$Description))]
```

```{r}
#Dropping original categorical variable and baseline (white) to avoid multicolinearity 
FINAL_ANALYSIS_DT <- FINAL_ANALYSIS_DT[, -c("Metro - Counties in metro areas of 1 million population or more","Description")]
```

```{r}
FINAL_ANALYSIS_DT <- FINAL_ANALYSIS_DT[!duplicated(DESYNPUF_ID)]
FINAL_ANALYSIS_DT <- FINAL_ANALYSIS_DT[,-c("DESYNPUF_ID")]
```

```{r}
# removing all observations with missing data
FINAL_ANALYSIS_DT <- na.omit(FINAL_ANALYSIS_DT)
```



# Exploratory data analysis

## Baseline Treatment Balance: Obesity v. Non-Obesity
```{r}
ggplot(FINAL_ANALYSIS_DT, aes(x = factor(obese), fill = factor(obese))) +
  geom_bar() +
  labs(title = "Treatment Balance",
       x = "Obese",
       y = "Count") +
  scale_fill_manual(values = c("grey", "maroon")) +
  theme_minimal()
```

## Baseline Outcome Balance: ER-Visit vs. Non-ER procedure
```{r}
ggplot(FINAL_ANALYSIS_DT, aes(x = factor(ER_visit), fill = factor(ER_visit))) +
  geom_bar() +
  labs(title = "Outcome Class Balance",
       x = "Obese",
       y = "Count") +
  scale_fill_manual(values = c("grey", "lightblue")) +
  theme_minimal()

```

```{r}
library(ggplot2)
table_matrix <- table(FINAL_ANALYSIS_DT$obese, FINAL_ANALYSIS_DT$ER_visit)
matrix_df <- as.data.frame(as.table(table_matrix))

ggplot(matrix_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), vjust = 1) +
  labs(title = "Baseline Population Overview: Obesity v. ER Visits",
       x = "Obese",
       y = "ER Visit",
       fill = "Frequency") +
  scale_fill_gradient(low = "grey", high = "maroon") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))


```


## Balancing Treatment Classes
```{r}
# Since this in a causal inference problem we won't balance the outcome variable
# However, it is important that our treatment groups are balanced
# To achieve this randomly downsampling majority (obese=1) group
nonobese_dt <- FINAL_ANALYSIS_DT[obese==0]
obese_dt <- FINAL_ANALYSIS_DT[obese==1]
```

```{r}
set.seed(12341)
obese_samp <- obese_dt[sample(nrow(obese_dt), nrow(nonobese_dt), replace = FALSE), ]
```

```{r}
FINAL_ANALYSIS_DT <- rbindlist(list(obese_samp, nonobese_dt),
                            use.names = TRUE)
```

```{r}
# rechecking Population obesity breakdown
sum(FINAL_ANALYSIS_DT$obese==1)
sum(FINAL_ANALYSIS_DT$obese==0)
```

```{r}
table_matrix <- table(FINAL_ANALYSIS_DT$obese, FINAL_ANALYSIS_DT$ER_visit)

matrix_df <- as.data.frame(as.table(table_matrix))

ggplot(matrix_df, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), vjust = 1) +
  labs(title = "Population Overview: Obesity v. ER Visits",
       x = "Obese",
       y = "ER Visit",
       fill = "Frequency") +
  scale_fill_gradient(low = "grey", high = "maroon") +
  theme_minimal()  +
  theme(axis.text.x = element_text(size = 12),  
        axis.text.y = element_text(size = 12))  
```


## Fitting preliminary logistic regression

```{r, results='hide'}
fit.1 <- glm(ER_visit ~ ., data = FINAL_ANALYSIS_DT, family = "binomial")
summary(fit.1)
```

```{r}
summary_fit.1 <- summary(fit.1)
variables_fit.1 <- as.data.table(rownames(summary_fit.1$coefficients))
coefficients_fit.1 <- as.data.table(summary_fit.1$coefficients)
coefficients_fit.1 <- coefficients_fit.1[, Sig := ifelse(`Pr(>|z|)` < 0.001, "***", 
                                                         ifelse(`Pr(>|z|)` < 0.01, "**",
                                                                ifelse(`Pr(>|z|)` < 0.05, "*", "")))]
abbreviated_covariates_fit.1 <- str_trunc(variables_fit.1, 150) 
summary_fit.1_dt <- cbind(variables_fit.1, coefficients_fit.1)
```

```{r}
kable(summary_fit.1_dt, digits = 4) %>% 
  kable_styling(font_size = 12, latex_options = "striped") %>% 
    add_header_above(c("Logistic Regression" = 6))
rm(variables_fit.1, coefficients_fit.1)
```


### Without estimating propensity scores being obese increases relative risk of ER visit ~25% and result is significant (p<.001)



# Causal Inference

## Approach 1: logisitic regression using propensity scores


```{r}
# Calculating standardized mean differences to decide which variables will be included in propensity scores (SMD>0.1 are considered imbalanced)
library(tableone)
vars <- setdiff(names(FINAL_ANALYSIS_DT), c("obese", "ER_visit"))

## Construct a table with "obese" as the treatment variable
tabUnmatched <- CreateTableOne(vars = vars, data = FINAL_ANALYSIS_DT, test = FALSE, 
                               strata = "obese")

## Show table with SMD
print(tabUnmatched, smd = TRUE)

```


```{r}
Tr <- cbind(FINAL_ANALYSIS_DT$obese)
Y <- cbind(FINAL_ANALYSIS_DT$ER_visit)
X <- cbind(FINAL_ANALYSIS_DT[, -c("ER_visit", 
                                  "obese"
                               )])
X <- as.matrix(X)
X2 <- cbind(FINAL_ANALYSIS_DT[, -c("obese", 
                               "ER_visit"
                               )])
X2 <- as.matrix(X2)
```

```{r, results='hide'}
# running first regression covariates on treatment
glm.treat <- glm(Tr ~ X, 
                 family = "binomial",
                 data = FINAL_ANALYSIS_DT)
summary(glm.treat)
```




```{r}
# Overall covariate values appear to be balanced among treatment groups (SMD<0.05)
# However, given the relatively high SMDs of kidney disease, alzheimer_dementia, 
# stroke_transient_ischemic_attack, and part_D_coverage, that they are significantly 
# correlated with the treatment (obesity), and a theoretical understanding that 
# they could be potential confounds, I will use these variables in the propensity score calculations
```

```{r, results='hide'}
# Estimating propensity scores
glm.treat <- glm(Tr ~ alzheimer_dementia + kidney_disease + stroke_transient_ischemic_attack + part_D_coverage - ER_visit,
                 family = "binomial",
                 data = FINAL_ANALYSIS_DT)
ps <- predict(glm.treat, type = "response")

# Calculate the inverse probability weights
inverse_weights <- ifelse(Tr == 1, 1 / ps, 1 / (1 - ps))

# Fit the outcome model using all variables, including the ones used for propensity score
fit.ps <- glm(Y ~ .- ER_visit , 
              family = "binomial", 
              weights = inverse_weights,
              data = FINAL_ANALYSIS_DT)
summary(fit.ps)

```

### Note data was already relatively balanced before estimating propensity score weights (SMD<0.05), so propensity scores had a negligible effect

```{r}
summary_fit.ps <- summary(fit.ps)
variables_fit.ps <- as.data.table(rownames(summary_fit.ps$coefficients))
coefficients_fit.ps <- as.data.table(summary_fit.ps$coefficients)
coefficients_fit.ps <- coefficients_fit.ps[, Sig := ifelse(`Pr(>|z|)` < 0.001, "***", 
                                                         ifelse(`Pr(>|z|)` < 0.01, "**",
                                                                ifelse(`Pr(>|z|)` < 0.05, "*", "")))]
abbreviated_covariates_fit.ps <- str_trunc(variables_fit.ps, 150) 
summary_fit.ps_dt <- cbind(variables_fit.ps, coefficients_fit.ps)
```

```{r}
kable(summary_fit.ps_dt, digits = 4) %>% 
  kable_styling(font_size = 12, latex_options = "striped") %>% 
    add_header_above(c("Logistic Regression" = 6))
rm(variables_fit.ps, coefficients_fit.ps)
```




## Approach 2: Causal LASSO 

```{r}
cv.treat <- cv.gamlr(X2,Tr, family = "binomial")
dhat <- predict(cv.treat$gamlr, X2, type = "response")
dhat <- drop(dhat)
```

```{r}
causal <- gamlr(cbind(Tr,dhat,X2),Y,
 family="binomial",free=2)
plot(causal)
```

```{r,  results='hide'}
coef(causal)
```


```{r}
causal_mat <- coef(causal)
causal_df <- as.data.frame(as.matrix(causal_mat))
kable(causal_df, digits = 4) %>% 
  kable_styling(font_size = 12, latex_options = "striped") %>% 
    add_header_above(c("Causal LASSO" = 2))
```





